A No-Go Theorem for $\psi$-Ontic Models

### 2022-06-10 Articles/2021-HarriganSpekkens/HS-ShortNote.tex
HarriganSpekkens: comments from Andrea

\documentclass[10pt,twocolumn, nofootinbib]{revtex4-2}
%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
%\documentclass[10pt,twocolumn,letterpaper]{article}

@@

\title{A no-go theorem for $\psi$-ontic models}
\author{Gabriele Carcassi}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Andrea Oldofredi}
\affiliation{Centre of Philosophy, University of Lisbon, Portugal}
\author{Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

\date{}

\begin{abstract}
In this note we give a simple argument as to why $\psi$-ontic models cannot be expected to reproduce quantum mechanics. We will see that, using information theoretic considerations, the lack of overlap of epistemic states requires all states to be orthogonal, which directly contradicts quantum theory. This argument is a much simplified version of the one we presented in a previous paper.
\end{abstract}

@@

\section{Introduction}

@@

\textbf{Non-overlapping distributions can only represent orthogonal states}. Now, suppose we have a $\psi$-ontic model. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping distributions over a space $\Lambda$, therefore eq. \ref{entropy_nonoverlap} applies. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
	H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq. \ref{entropy_mixed}, it follows that $p$ must be zero. That is, we must have that
\begin{equation}\label{orthogonal}
	 \langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

The non-overlapping assumption built into the $\psi$-ontic model necessarily implies that all pure states are orthogonal. Since this is not true in quantum mechanics, any $\psi$-ontic models will fail to reproduce the results of quantum information, quantum statistical mechanics and, therefore, quantum theory in general.

In retrospect this should not be surprising for two reasons. First of all, the case where all states are orthogonal is exactly the case for which all observables commute, the classical case. Therefore we are simply finding that the implicit use of classical probability in $\psi$-ontic models pushes us to the classical case. Secondly, we already know that quantum information theory has features that are not reproducible in classical information theory, and that is exactly what makes it such an exciting new field.

@@

As $\rho$ is in a diagonal form, the entropy is given by eq.~\ref{entropy}.

\bibliography{bibliography}

\end{document}

>>>>

\documentclass[10pt,twocolumn, nofootinbib]{revtex4-1}
%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
%\documentclass[10pt,twocolumn,letterpaper]{article}

@@

\title{A No-Go Theorem for $\psi$-Ontic Models}
\author{Gabriele Carcassi}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Andrea Oldofredi}
\affiliation{Centre of Philosophy, University of Lisbon, Portugal}
\author{Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

\date{9 June 2022}

\begin{abstract}
In this note we give a simple no-go theorem that explains why $\psi$-ontic models cannot be expected to reproduce quantum mechanics. We will see that, using information theoretic considerations, the lack of overlap of epistemic state requires all states to be orthogonal, which openly contradicts quantum theory. This argument is a simplified version of the one we presented in a \href{https://arxiv.org/pdf/2201.11842.pdf}{previous paper}.
\end{abstract}

@@

\section{Introduction}
In 2010 Harrigan and Spekkens (HS) proposed a formal classification to categorize the nature of the quantum state, i.e.\ to establish whether in a certain model $\psi$ corresponds to a real property instantiated by a quantum object, in which case the model is called $\psi$-ontic, or to some observer information, making it $\psi$-epistemic \cite{Harrigan:2010}. Although the original aim of this classification was to clarify Einstein's view about quantum mechanics (QM), the HS framework has been widely employed not only to categorize different interpretations of QM, but also to argue what formulations of the theory are admissible (\cite{PBR:2012, Leifer:2013, Leifer:2017, Branciard:2014, Hermens:2021, Wood:2015, Ringbauer:2015, Mazurek:2016, Bartlett:2012}; cf.\ \cite{Oldofredi:2020b, Lewis:2012, Ladyman:2021} for critical discussions).

Given the influence of the HS framework in quantum foundations, and in particular the prominent role played by $\psi-$ontic models in several arguments supporting the reality of the quantum state---many of them built upon \cite{PBR:2012}---we re-examine the definitions and features of such models in order to understand whether they constitute a sound basis from which to draw conclusions on the nature of $\psi$. Our analysis suggests that they are not.

Referring to this, in this note we provide a no-go theorem for $\psi-$ontic models which is a simplified version of the main result contained in our previous paper \cite{Carcassi:2022}. More precisely, here we show that, using information theoretic considerations, $\psi-$ontic models cannot reproduce QM since the lack of overlap of epistemic state requires all states to be orthogonal, which openly contradicts quantum theory. Therefore, we conclude, they should not be employed in order to classify quantum models.\footnote{We assume that the reader is familiar with HS definitions.}

@@

\textbf{Theorem:} \emph{Non-overlapping distributions can only represent orthogonal states}.

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping distributions over a space $\Lambda$, therefore eq. \ref{entropy_nonoverlap} applies. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
	H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.\ \ref{entropy_mixed}, it follows that $p$ must be zero. That is, we must have that
\begin{equation}\label{orthogonal}
	 \langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

Hence, the non-overlapping assumption built into the $\psi$-ontic model necessarily implies that all pure states are orthogonal. Since this is not true in quantum mechanics, any $\psi$-ontic model will fail to reproduce the results of quantum information, quantum statistical mechanics and, therefore, quantum theory in general.

In retrospect this should not be surprising for two reasons. First of all, the case where all states are orthogonal corresponds exactly to the case in which all observables commute, i.e.	\ the classical case. Therefore, we are simply finding that the implicit use of classical probability in $\psi$-ontic models pushes us to classical interpretation of quantum models. Secondly, we already know that quantum information theory has features that are not reproducible in classical information theory, and that is exactly what makes it such an exciting new field.

@@

As $\rho$ is in a diagonal form, the entropy is given by eq.~\ref{entropy}.\end{document}
